---
title: "Space-Filling Designs for Computer Experiments"
date: today
lang: en
format:
  html:
    toc: true
    toc-title: "Contents"
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: flatly
    highlight-style: tango
    code-fold: show
    code-tools: true
    smooth-scroll: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r}
#| label: packages
#| code-fold: true
#| code-summary: "Load packages & theme"

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(lhs)          # Latin Hypercube Sampling
library(randtoolbox)  # Sobol, Halton sequences
library(DiceDesign)   # maximin / minimax / optimisation
library(DiceKriging)  # GP emulator for sequential section

col1 <- "#e94560"
col2 <- "#0f3460"
col3 <- "#f5a623"
col4 <- "#16a085"
col5 <- "#533483"

theme_sfd <- theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(color = "grey40", size = 10),
    strip.text      = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

------------------------------------------------------------------------

# Why Space-Filling Designs?

## The Core Problem

When running a **computer experiment** (CFD simulation, finite-element model, agent-based model, climate code…), every function evaluation is expensive. Given a fixed budget of $n$ runs, *where* should we place those runs in the input space $\mathcal{X} \subseteq \mathbb{R}^d$?

Classical design of experiments (full factorial, central composite, Box–Behnken) was built for **physical** experiments that suffer from random noise: replication matters there. Computer codes are **deterministic** — the same input always returns the same output — so replication is wasteful. What we want instead is **coverage**: spread the $n$ points as uniformly as possible over $\mathcal{X}$, so that no region is left unexplored.

::: callout-note
A **space-filling design** is a set of $n$ points that cover the input space as uniformly as possible, ensuring good global approximation with the minimum number of evaluations.
:::

## What Makes a Design "Good"?

Several complementary criteria exist:

| Criterion | Intuition | Design type |
|----|----|----|
| **Maximin distance** | Maximise the minimum pairwise distance | Points as far apart as possible |
| **Minimax distance** | Minimise the maximum distance to the nearest design point | No region left uncovered |
| **Discrepancy** | Minimise deviation from a perfectly uniform distribution | Low-discrepancy sequences |
| **Projective uniformity** | Uniform in every 1D and 2D projection | Latin Hypercube |
| **Entropy** | Maximise information content (GP-based) | Maximum entropy designs |

------------------------------------------------------------------------

# Simple Random Sampling: the Baseline

Before looking at sophisticated designs, let's see why naive **random sampling** often fails, especially in low-$n$ regimes.

```{r}
#| label: random-sampling
#| fig-height: 5

set.seed(42)
n <- 20

df_random <- data.frame(x1 = runif(n), x2 = runif(n))

nn_dist <- function(pts) {
  D <- as.matrix(dist(pts)); diag(D) <- Inf
  apply(D, 1, min)
}

cat("Random sampling — min pairwise distance:",
    round(min(dist(df_random)), 4), "\n")

ggplot(df_random, aes(x1, x2)) +
  geom_point(color = col1, size = 3.5, alpha = 0.85) +
  geom_vline(xintercept = seq(0, 1, by = 0.2), linetype = "dotted", color = "grey80") +
  geom_hline(yintercept = seq(0, 1, by = 0.2), linetype = "dotted", color = "grey80") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title    = "Random Sampling (n = 20)",
    subtitle = "Clusters and large empty regions are common",
    x = expression(x[1]), y = expression(x[2])
  ) + theme_sfd
```

With random sampling, **clusters** appear in some regions while others remain empty — exactly what we want to avoid when building a surrogate model.

------------------------------------------------------------------------

# Latin Hypercube Sampling (LHS)

## Principle

A **Latin Hypercube Sample** of size $n$ in $d$ dimensions is constructed by:

1.  Dividing each axis into $n$ equal strata $[0,\tfrac{1}{n}),\,[\tfrac{1}{n},\tfrac{2}{n}),\ldots,[\tfrac{n-1}{n},1]$.
2.  Placing exactly **one point per stratum** on each axis (the "Latin" constraint).
3.  Combining strata across dimensions using a random permutation.

This guarantees **perfect 1D projective uniformity**: every stratum contains exactly one point on any single-axis projection.

$$X_{ij} = \frac{\pi_j(i) - U_{ij}}{n}, \quad U_{ij} \sim \mathcal{U}(0,1), \quad \pi_j \text{ random permutation of } \{1,\ldots,n\}$$

```{r}
#| label: lhs-basic
#| fig-height: 5

set.seed(42)
n <- 20

X_lhs_rand <- randomLHS(n, 2)
X_lhs_opt  <- optimumLHS(n, 2, maxSweeps = 4, eps = 0.1)

df_lhs <- bind_rows(
  data.frame(x1 = X_lhs_rand[,1], x2 = X_lhs_rand[,2], type = "Random LHS"),
  data.frame(x1 = X_lhs_opt[,1],  x2 = X_lhs_opt[,2],  type = "Optimised LHS")
)

make_lhs_plot <- function(df, col, ttl) {
  ggplot(df, aes(x1, x2)) +
    geom_vline(xintercept = seq(0, 1, by = 1/n), color = "grey85", linewidth = 0.4) +
    geom_hline(yintercept = seq(0, 1, by = 1/n), color = "grey85", linewidth = 0.4) +
    geom_point(color = col, size = 3.5, alpha = 0.9) +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(title = ttl, x = expression(x[1]), y = expression(x[2])) +
    theme_sfd
}

make_lhs_plot(df_lhs |> filter(type == "Random LHS"),    col1, "Random LHS") +
make_lhs_plot(df_lhs |> filter(type == "Optimised LHS"), col2, "Optimised LHS (maximin)") +
plot_annotation(
  title    = "Latin Hypercube Sampling — n = 20 points in [0,1]\u00b2",
  subtitle = "Grid lines show the n \u00d7 n strata. Each row and column contains exactly one point.",
  theme    = theme(plot.title = element_text(face = "bold", size = 13))
)
```

## 1D Projective Uniformity

The defining property of LHS: **perfect uniformity on every axis**.

```{r}
#| label: lhs-projection
#| fig-height: 7

set.seed(42)
X_lhs  <- randomLHS(n, 2)
X_rand <- matrix(runif(n * 2), ncol = 2)

make_proj <- function(X, label, col) {
  df <- data.frame(x1 = X[,1], x2 = X[,2])

  p_main <- ggplot(df, aes(x1, x2)) +
    geom_point(color = col, size = 2.5) +
    coord_fixed(xlim = c(0,1), ylim = c(0,1)) +
    labs(title = label, x = expression(x[1]), y = expression(x[2])) +
    theme_sfd

  p_x1 <- ggplot(df, aes(x = x1, y = 0)) +
    geom_point(color = col, size = 2, shape = 17) +
    geom_vline(xintercept = seq(0, 1, by = 1/n), color = "grey80", linewidth = 0.3) +
    xlim(0, 1) + ylim(-0.5, 0.5) +
    labs(x = expression(x[1]), y = NULL) + theme_sfd +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

  list(main = p_main, x1 = p_x1)
}

pr_rand <- make_proj(X_rand, "Random Sampling", col1)
pr_lhs  <- make_proj(X_lhs,  "LHS",             col2)

(pr_rand$main + pr_lhs$main) /
(pr_rand$x1   + pr_lhs$x1) +
plot_annotation(
  title    = "Projective Uniformity: LHS vs Random Sampling",
  subtitle = "Bottom row: projection onto x\u2081 axis. LHS fills each stratum exactly once.",
  theme    = theme(plot.title = element_text(face = "bold"))
)
```

## LHS Variants in the `lhs` Package

```{r}
#| label: lhs-variants
#| fig-height: 6

set.seed(123)

designs <- list(
  "randomLHS"  = randomLHS(n, 2),
  "maximinLHS" = maximinLHS(n, 2),
  "optimumLHS" = optimumLHS(n, 2, maxSweeps = 4, eps = 0.1),
  "augmentLHS" = augmentLHS(randomLHS(10, 2), 10)
)

df_variants <- lapply(names(designs), function(nm) {
  data.frame(x1 = designs[[nm]][,1], x2 = designs[[nm]][,2], design = nm)
}) |> bind_rows()

colors_v <- c(randomLHS = col1, maximinLHS = col2,
               optimumLHS = col3, augmentLHS = col4)

ggplot(df_variants, aes(x1, x2, color = design)) +
  geom_vline(xintercept = seq(0,1,by=1/n), color="grey90", linewidth=0.3) +
  geom_hline(yintercept = seq(0,1,by=1/n), color="grey90", linewidth=0.3) +
  geom_point(size = 3) +
  scale_color_manual(values = colors_v) +
  facet_wrap(~design, ncol = 2) +
  coord_fixed(xlim=c(0,1), ylim=c(0,1)) +
  labs(
    title    = "LHS Variants \u2014 n = 20 in [0,1]\u00b2",
    subtitle = "Each variant optimises a different quality criterion",
    x = expression(x[1]), y = expression(x[2])
  ) + theme_sfd + theme(legend.position = "none")
```

------------------------------------------------------------------------

# Maximin and Minimax Designs

## Definitions

**Maximin design** — maximise the minimum pairwise distance:

$$\text{Maximin:} \quad \max_{\mathbf{X}} \min_{i \neq j} d(\mathbf{x}_i, \mathbf{x}_j)$$

This ensures no two design points are unnecessarily close — points are "as far apart as possible".

**Minimax design** — minimise the maximum distance from any point in $\mathcal{X}$ to the nearest design point:

$$\text{Minimax:} \quad \min_{\mathbf{X}} \max_{\mathbf{x} \in \mathcal{X}} \min_i d(\mathbf{x}, \mathbf{x}_i)$$

This ensures no region is left too far from a design point — a coverage guarantee.

```{r}
#| label: maximin-minimax
#| fig-height: 5

set.seed(42)
n_mm <- 15

mm_result  <- dmaxDesign(n_mm, 2, 200, "unif", 100)
X_maximin  <- mm_result$design

phi_result <- phiPDesign(n_mm, 2, 50)
X_minimax  <- phi_result$design

dist_stats <- function(X, label) {
  D <- as.matrix(dist(X)); diag(D) <- Inf
  mins <- apply(D, 1, min)
  data.frame(design = label, min_dist = min(mins),
             mean_min = mean(mins), n = nrow(X))
}

stats_mm <- bind_rows(
  dist_stats(X_maximin, "Maximin"),
  dist_stats(X_minimax, "PhiP (minimax proxy)")
)
print(stats_mm)

coverage_map <- function(X, res = 50) {
  grid  <- expand.grid(x1 = seq(0,1,length.out=res),
                       x2 = seq(0,1,length.out=res))
  D     <- as.matrix(dist(rbind(as.matrix(grid), X)))
  n_g   <- nrow(grid)
  D_sub <- D[1:n_g, (n_g+1):(n_g+nrow(X))]
  grid$coverage <- apply(D_sub, 1, min)
  grid
}

cov_maximin <- coverage_map(X_maximin)
cov_minimax <- coverage_map(X_minimax)

make_cov_plot2 <- function(cov_df, X, title_txt) {
  ggplot(cov_df, aes(x1, x2, fill = coverage)) +
    geom_tile() +
    geom_point(data = as.data.frame(X), aes(V1, V2),
               inherit.aes = FALSE, color = "white", size = 2.5, shape = 16) +
    scale_fill_viridis_c(option = "magma", direction = -1,
                          name = "Dist. to\nnearest\npoint") +
    coord_fixed() +
    labs(title = title_txt, x = expression(x[1]), y = expression(x[2])) +
    theme_sfd
}

make_cov_plot2(cov_maximin, X_maximin, "Maximin Design \u2014 Coverage Map") +
make_cov_plot2(cov_minimax, X_minimax, "PhiP (Minimax Proxy) \u2014 Coverage Map") +
plot_annotation(
  title    = "Coverage maps: dark = well covered, bright = far from any design point",
  subtitle = paste0("n = ", n_mm, " points. Minimax minimises the brightest regions."),
  theme    = theme(plot.title = element_text(face = "bold"))
)
```

## Comparing Designs: Nearest-Neighbour Distances

```{r}
#| label: design-comparison-dist
#| fig-height: 5

set.seed(42)
n_c <- 20

designs_c <- list(
  "Random"        = matrix(runif(n_c*2), ncol=2),
  "Random LHS"    = randomLHS(n_c, 2),
  "Optimised LHS" = optimumLHS(n_c, 2, maxSweeps=4, eps=0.1),
  "Maximin"       = dmaxDesign(n_c, 2, 200, "unif", 100)$design
)

df_dist <- lapply(names(designs_c), function(nm) {
  X <- designs_c[[nm]]
  D <- as.matrix(dist(X)); diag(D) <- Inf
  data.frame(min_d = apply(D, 1, min), design = nm)
}) |> bind_rows()

df_dist$design <- factor(df_dist$design,
  levels = c("Random","Random LHS","Optimised LHS","Maximin"))

ggplot(df_dist, aes(x = design, y = min_d, fill = design)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  scale_fill_manual(values = c(col1, col2, col3, col4)) +
  labs(
    title    = "Distribution of Nearest-Neighbour Distances by Design Type",
    subtitle = "Higher minimum distance = better space coverage",
    x = NULL, y = "Min pairwise distance"
  ) + theme_sfd + theme(legend.position = "none")
```

------------------------------------------------------------------------

# Low-Discrepancy Sequences

## What is Discrepancy?

**Discrepancy** measures how far a point set deviates from perfect uniformity. For a set of $n$ points in $[0,1]^d$, the **star discrepancy** is:

$$D_n^* = \sup_{J \subseteq [0,1]^d} \left| \frac{\#\{i : \mathbf{x}_i \in J\}}{n} - \mathrm{Vol}(J) \right|$$

Low-discrepancy sequences achieve $D_n^* = \mathcal{O}\!\left((\log n)^d / n\right)$, versus $\mathcal{O}(n^{-1/2})$ for random sampling.

## Halton Sequence

The Halton sequence uses different **prime bases** for each dimension. For dimension $k$ with base $b_k$:

$$x_{n,k} = \sum_{j=0}^\infty d_j \, b_k^{-(j+1)}, \quad \text{where } n = \sum_j d_j b_k^j$$

```{r}
#| label: halton
#| fig-height: 5

n_ld     <- 64
X_halton <- halton(n_ld, dim = 2)
X_rand_ld <- matrix(runif(n_ld * 2), ncol = 2)

df_ld <- bind_rows(
  data.frame(x1=X_rand_ld[,1], x2=X_rand_ld[,2], type="Random"),
  data.frame(x1=X_halton[,1],  x2=X_halton[,2],  type="Halton Sequence")
)

make_ld_plot <- function(df, col, ttl) {
  ggplot(df, aes(x1, x2)) +
    geom_point(color = col, size = 1.8, alpha = 0.8) +
    coord_fixed(xlim=c(0,1), ylim=c(0,1)) +
    labs(title=ttl, x=expression(x[1]), y=expression(x[2])) +
    theme_sfd
}

make_ld_plot(df_ld |> filter(type=="Random"),          col1, "Random (n=64)") +
make_ld_plot(df_ld |> filter(type=="Halton Sequence"), col2, "Halton Sequence (n=64)") +
plot_annotation(
  title    = "Halton vs Random \u2014 n = 64 points",
  subtitle = "Halton avoids clusters through deterministic base-prime construction",
  theme    = theme(plot.title=element_text(face="bold"))
)
```

## Sobol Sequence

The **Sobol sequence** uses base-2 arithmetic and direction numbers to achieve excellent uniformity, especially in higher dimensions. It is the workhorse of quasi-Monte Carlo integration and design.

```{r}
#| label: sobol
#| fig-height: 5

n_sobol <- 64
X_sobol_rand <- sobol(n_sobol, dim=2, scrambling=0)
X_sobol_scr  <- sobol(n_sobol, dim=2, scrambling=1, seed=42)

df_sobol <- bind_rows(
  data.frame(x1=X_sobol_rand[,1], x2=X_sobol_rand[,2], type="Sobol (unscrambled)"),
  data.frame(x1=X_sobol_scr[,1],  x2=X_sobol_scr[,2],  type="Sobol (Owen scrambled)")
)

make_ld_plot(df_sobol |> filter(type=="Sobol (unscrambled)"),    col3, "Sobol \u2014 Unscrambled") +
make_ld_plot(df_sobol |> filter(type=="Sobol (Owen scrambled)"), col4, "Sobol \u2014 Owen Scrambled") +
plot_annotation(
  title    = "Sobol Sequences \u2014 n = 64 points",
  subtitle = "Scrambling randomises the sequence while preserving low discrepancy",
  theme    = theme(plot.title=element_text(face="bold"))
)
```

## Nested Property: Sobol Sequences Scale Well

A key advantage of Sobol sequences: the first $2^k$ points are already a good design. You can **add more points** without redoing the whole experiment.

```{r}
#| label: sobol-nested
#| fig-height: 4

sizes <- c(8, 16, 32, 64)

df_nested <- lapply(sizes, function(n_i) {
  X <- sobol(n_i, dim=2, scrambling=1, seed=42)
  data.frame(x1=X[,1], x2=X[,2],
             n_label = paste0("n = ", n_i),
             is_new  = (seq_len(n_i)) > (n_i / 2))
}) |> bind_rows()

df_nested$n_label <- factor(df_nested$n_label, levels=paste0("n = ", sizes))

ggplot(df_nested, aes(x1, x2, color=is_new, size=is_new)) +
  geom_point(alpha=0.9) +
  scale_color_manual(values=c("FALSE"=col2,"TRUE"=col1),
                      labels=c("FALSE"="Previous points","TRUE"="New points")) +
  scale_size_manual(values=c("FALSE"=1.5,"TRUE"=2.5), guide="none") +
  facet_wrap(~n_label, ncol=4) +
  coord_fixed(xlim=c(0,1), ylim=c(0,1)) +
  labs(
    title    = "Nested Structure of Sobol Sequences",
    subtitle = "Red points are new additions. Previous points remain valid \u2014 no redesign needed.",
    x=expression(x[1]), y=expression(x[2]), color=NULL
  ) + theme_sfd
```

------------------------------------------------------------------------

# Side-by-Side Comparison of All Designs

```{r}
#| label: all-designs
#| fig-height: 8

set.seed(42)
n_all <- 32

X_list <- list(
  "Random"       = matrix(runif(n_all*2), ncol=2),
  "Random LHS"   = randomLHS(n_all, 2),
  "Optim. LHS"   = optimumLHS(n_all, 2, maxSweeps=4, eps=0.1),
  "Maximin"      = dmaxDesign(n_all, 2, 200, "unif", 100)$design,
  "Halton"       = halton(n_all, dim=2),
  "Sobol"        = sobol(n_all, dim=2, scrambling=1, seed=42)
)

df_all <- lapply(names(X_list), function(nm) {
  data.frame(x1=X_list[[nm]][,1], x2=X_list[[nm]][,2], design=nm)
}) |> bind_rows()

df_all$design <- factor(df_all$design, levels=names(X_list))
pal6 <- c(col1, col2, col3, col4, col5, "#c0392b")

ggplot(df_all, aes(x1, x2, color=design)) +
  geom_point(size=2.2, alpha=0.9) +
  scale_color_manual(values=pal6) +
  facet_wrap(~design, ncol=3) +
  coord_fixed(xlim=c(0,1), ylim=c(0,1)) +
  labs(
    title = "All Space-Filling Designs \u2014 n = 32 points",
    x=expression(x[1]), y=expression(x[2])
  ) + theme_sfd + theme(legend.position="none")
```

## Quantitative Comparison

```{r}
#| label: design-metrics

design_metrics <- function(X, name) {
  D    <- as.matrix(dist(X)); diag(D) <- Inf
  mins <- apply(D, 1, min)
  disc <- tryCatch(
    unname(discrepancyCriteria(X)$DiscrStarStar),
    error = function(e) NA
  )
  data.frame(Design      = name,
             Min_dist    = round(min(mins), 4),
             Mean_min    = round(mean(mins), 4),
             Discrepancy = round(disc, 5))
}

df_metrics <- lapply(names(X_list), function(nm) {
  design_metrics(X_list[[nm]], nm)
}) |> bind_rows()

knitr::kable(
  df_metrics, align = "lrrr",
  col.names = c("Design","Min distance","Mean min dist","Star discrepancy"),
  caption   = paste0("Design quality metrics \u2014 n=", n_all, " points, d=2")
)
```

------------------------------------------------------------------------

# Application: Tree Placement in a Maize Field

## Problem Setting

Agroforestry researchers want to place **sensor trees** in a maize field to monitor micro-climate variables (light, humidity, temperature). Each tree carries sensors and acts as a monitoring station. The goals are:

-   Cover the field uniformly so that **any point is close to at least one sensor tree**.
-   Avoid clusters (redundant sensors).
-   Fit a **GPS-defined irregular boundary** (roads, water canal on one side).

This is a **space-filling design problem in a non-rectangular domain**.

```{r}
#| label: field-setup
#| fig-height: 5

field_poly <- data.frame(
  x = c(0, 200, 200, 150, 100,  0,   0),
  y = c(0,   0, 100, 120, 120, 120,  0)
)

point_in_poly <- function(px, py, poly) {
  np  <- nrow(poly) - 1
  inside <- logical(length(px))
  for (k in seq_along(px)) {
    xp <- px[k]; yp <- py[k]; j <- np; ins <- FALSE
    for (i in 1:np) {
      xi <- poly$x[i]; yi <- poly$y[i]
      xj <- poly$x[j]; yj <- poly$y[j]
      if (((yi > yp) != (yj > yp)) &&
          (xp < (xj-xi)*(yp-yi)/(yj-yi)+xi)) ins <- !ins
      j <- i
    }
    inside[k] <- ins
  }
  inside
}

ggplot(field_poly, aes(x, y)) +
  geom_polygon(fill="#4CAF50", alpha=0.25, color=col2, linewidth=1) +
  annotate("text", x=100, y=55, label="Maize Field\n200m \u00d7 ~120m",
           size=5, color=col2, fontface="bold") +
  annotate("segment", x=150, xend=200, y=120, yend=100,
           color="#2196F3", linewidth=2) +
  annotate("text", x=177, y=115, label="Canal", color="#2196F3",
           size=3.5, angle=-20) +
  annotate("segment", x=0, xend=200, y=0, yend=0,
           color="#795548", linewidth=2.5) +
  annotate("text", x=100, y=-9, label="Road (South)", color="#795548", size=3.5) +
  scale_x_continuous(breaks=seq(0,200,50), labels=paste0(seq(0,200,50),"m")) +
  scale_y_continuous(breaks=seq(0,120,30), labels=paste0(seq(0,120,30),"m")) +
  coord_fixed() +
  labs(title="Maize Field Layout",
       subtitle="Irregular boundary due to a canal cutting the SE corner",
       x="East-West (m)", y="North-South (m)") +
  theme_sfd
```

## Generating Points Inside the Irregular Field

```{r}
#| label: field-designs
#| fig-height: 9
#| cache: true

set.seed(42)

sample_in_field <- function(n_target, method="sobol", seed=42) {
  batch <- n_target * 6
  if (method == "sobol") {
    raw   <- sobol(batch, dim=2, scrambling=1, seed=seed)
    cands <- data.frame(x=raw[,1]*200, y=raw[,2]*120)
  } else if (method == "lhs") {
    set.seed(seed)
    raw   <- randomLHS(batch, 2)
    cands <- data.frame(x=raw[,1]*200, y=raw[,2]*120)
  } else {
    set.seed(seed)
    cands <- data.frame(x=runif(batch,0,200), y=runif(batch,0,120))
  }
  mask <- point_in_poly(cands$x, cands$y, field_poly)
  cands[mask, ][seq_len(n_target), ]
}

n_trees      <- 20
trees_random <- sample_in_field(n_trees, "random", seed=42)
trees_lhs    <- sample_in_field(n_trees, "lhs",    seed=42)
trees_sobol  <- sample_in_field(n_trees, "sobol",  seed=42)

maximin_field <- function(pts_init, n_iter=500) {
  pts <- as.matrix(pts_init)
  for (iter in seq_len(n_iter)) {
    D <- as.matrix(dist(pts)); diag(D) <- Inf
    idx <- which.min(apply(D, 1, min))
    for (try in seq_len(50)) {
      cand <- c(runif(1,0,200), runif(1,0,120))
      if (point_in_poly(cand[1], cand[2], field_poly)) {
        pts_new       <- pts; pts_new[idx,] <- cand
        D_new         <- as.matrix(dist(pts_new)); diag(D_new) <- Inf
        if (min(D_new) >= min(D)) { pts <- pts_new; break }
      }
    }
  }
  as.data.frame(pts)
}

set.seed(42)
trees_maximin        <- maximin_field(trees_sobol)
names(trees_maximin) <- c("x","y")

coverage_field <- function(pts, res=60) {
  grid <- expand.grid(x=seq(0,200,length.out=res),
                      y=seq(0,120,length.out=res))
  mask <- point_in_poly(grid$x, grid$y, field_poly)
  D    <- as.matrix(dist(rbind(as.matrix(grid[,1:2]), as.matrix(pts))))
  n_g  <- nrow(grid)
  D_sub <- D[seq_len(n_g), (n_g+1):(n_g+nrow(pts))]
  grid$coverage         <- apply(D_sub, 1, min)
  grid$coverage[!mask]  <- NA
  grid
}

cov_random  <- coverage_field(trees_random)
cov_lhs     <- coverage_field(trees_lhs)
cov_sobol   <- coverage_field(trees_sobol)
cov_maximin <- coverage_field(trees_maximin)

make_field_plot <- function(cov_df, pts, title_txt) {
  ggplot() +
    geom_tile(data=cov_df, aes(x,y,fill=coverage), na.rm=TRUE) +
    geom_polygon(data=field_poly, aes(x,y), fill=NA, color="grey30", linewidth=0.7) +
    geom_point(data=pts, aes(x,y), color="white", size=3.5, shape=16) +
    geom_point(data=pts, aes(x,y), color=col2,    size=2,   shape=16) +
    scale_fill_viridis_c(option="magma", direction=-1,
                          name="Dist.(m)", na.value="white", limits=c(0,NA)) +
    coord_fixed() +
    labs(title=title_txt, x="East-West (m)", y="North-South (m)") +
    theme_sfd + theme(legend.position="right")
}

(make_field_plot(cov_random,  trees_random,  "Random placement") +
 make_field_plot(cov_lhs,     trees_lhs,     "LHS placement")) /
(make_field_plot(cov_sobol,   trees_sobol,   "Sobol placement") +
 make_field_plot(cov_maximin, trees_maximin, "Maximin placement")) +
plot_annotation(
  title    = "Tree Placement in Maize Field \u2014 Coverage Maps (n = 20 sensor trees)",
  subtitle = "Dark = close to a sensor tree (well monitored). Bright = poorly covered zones.",
  theme    = theme(plot.title=element_text(face="bold", size=13))
)
```

## Coverage Statistics

```{r}
#| label: field-stats

field_stats <- function(cov_df, pts, name) {
  vals <- na.omit(cov_df$coverage)
  D    <- as.matrix(dist(as.matrix(pts))); diag(D) <- Inf
  data.frame(
    Design        = name,
    Max_coverage  = round(max(vals), 1),
    Mean_coverage = round(mean(vals), 1),
    Min_pairwise  = round(min(D), 1)
  )
}

df_fstats <- bind_rows(
  field_stats(cov_random,  trees_random,  "Random"),
  field_stats(cov_lhs,     trees_lhs,     "LHS"),
  field_stats(cov_sobol,   trees_sobol,   "Sobol"),
  field_stats(cov_maximin, trees_maximin, "Maximin")
)

knitr::kable(
  df_fstats, align="lrrr",
  col.names = c("Design","Max dist. to tree (m)","Mean dist. (m)","Min pairwise dist. (m)"),
  caption   = "Field coverage statistics \u2014 20 sensor trees"
)
```

## Final Recommendation: Maximin Design

```{r}
#| label: field-final
#| fig-height: 5.5

ggplot() +
  geom_tile(data=cov_maximin, aes(x,y,fill=coverage), na.rm=TRUE, alpha=0.75) +
  scale_fill_gradient(low="#c8e6c9", high="#ff8f00",
                       name="Dist. to\nnearest tree (m)") +
  geom_polygon(data=field_poly, aes(x,y), fill=NA, color=col2, linewidth=1) +
  geom_point(data=trees_maximin, aes(x,y), color="white",   size=7,   shape=16) +
  geom_point(data=trees_maximin, aes(x,y), color="#2e7d32", size=5.5, shape=17) +
  geom_text(data = trees_maximin |> mutate(id=row_number()),
            aes(x, y, label=id), color="white", size=2.2, fontface="bold") +
  annotate("text", x=170, y=8,   label="Road",  color="#795548", size=3, fontface="italic") +
  annotate("text", x=178, y=112, label="Canal", color="#1565c0", size=3,
           fontface="italic", angle=-20) +
  coord_fixed() +
  scale_x_continuous(breaks=seq(0,200,50), labels=paste0(seq(0,200,50),"m")) +
  scale_y_continuous(breaks=seq(0,120,30), labels=paste0(seq(0,120,30),"m")) +
  labs(
    title    = "Optimal Sensor Tree Placement \u2014 Maximin Design",
    subtitle = paste0("20 sensor trees (green triangles) to maximise field coverage.\n",
                      "Colour gradient: distance to nearest sensor tree."),
    x="East-West (m)", y="North-South (m)"
  ) + theme_sfd
```

------------------------------------------------------------------------

# Sequential Design with an Acquisition Function

## Motivation: Goal-Oriented Exploration

A space-filling design treats all regions equally. But often we have a **specific objective** — find the maximum of $f$, locate a contour $f(\mathbf{x}) = c$, or minimise a response. In this case, we should invest more simulation budget **where it matters most**.

**Sequential design** (also called *active learning* or *Bayesian optimisation*) works as follows:

::: {.callout-tip title="Sequential Design Algorithm"}
1.  Start with a small initial design (e.g. LHS with $n_0$ points).
2.  Fit a **Gaussian Process** emulator on the current observations.
3.  Use an **acquisition function** $\alpha(\mathbf{x})$ to select the next point.
4.  Evaluate the expensive simulator at the chosen point.
5.  Update the GP and repeat until budget is exhausted.
:::

## Acquisition Functions

| Function | Formula | Goal |
|----|----|----|
| **Expected Improvement (EI)** | $\mathbb{E}[\max(\tau - f(\mathbf{x}), 0)]$ | Minimisation |
| **Upper Confidence Bound (UCB)** | $\mu(\mathbf{x}) + \kappa\,\sigma(\mathbf{x})$ | Maximisation |
| **Probability of Improvement (PI)** | $\Phi\!\left(\frac{\tau-\mu(\mathbf{x})}{\sigma(\mathbf{x})}\right)$ | Minimisation |
| **Integrated Variance** | $\int \sigma^2(\mathbf{x}\mid\mathbf{x}^+)\,d\mathbf{x}$ | Pure exploration |
| **Max Variance** | $\sigma^2(\mathbf{x})$ | Pure exploration |

## Case Study: Finding the Minimum of a Costly Function

We minimise the **Goldstein–Price function** (a classic multi-modal benchmark) using Sequential Bayesian Optimisation with Expected Improvement.

```{r}
#| label: seq-design-setup
#| fig-height: 5

goldstein_price <- function(x1, x2) {
  x1 <- 4*x1 - 2; x2 <- 4*x2 - 2
  A  <- 1 + (x1+x2+1)^2*(19-14*x1+3*x1^2-14*x2+6*x1*x2+3*x2^2)
  B  <- 30 + (2*x1-3*x2)^2*(18-32*x1+12*x1^2+48*x2-36*x1*x2+27*x2^2)
  log(A * B)
}

grid_gp        <- expand.grid(x1=seq(0,1,length.out=80), x2=seq(0,1,length.out=80))
grid_gp$z_true <- mapply(goldstein_price, grid_gp$x1, grid_gp$x2)

true_min_idx <- which.min(grid_gp$z_true)
true_min_x   <- as.numeric(grid_gp[true_min_idx, 1:2])
true_min_z   <- grid_gp$z_true[true_min_idx]

ggplot(grid_gp, aes(x1, x2, fill=z_true)) +
  geom_tile() +
  geom_point(data=data.frame(x1=true_min_x[1], x2=true_min_x[2]),
             aes(x1, x2), color="white", size=4, shape=8) +
  scale_fill_viridis_c(option="plasma", name="f(x)") +
  coord_fixed() +
  labs(
    title    = "Goldstein\u2013Price Function (log-transformed)",
    subtitle = "White star: global minimum. Goal: find it with as few evaluations as possible.",
    x=expression(x[1]), y=expression(x[2])
  ) + theme_sfd
```

## Acquisition Function: Expected Improvement

$$\mathrm{EI}(\mathbf{x}) = (\tau - \mu(\mathbf{x}))\,\Phi\!\left(\frac{\tau - \mu(\mathbf{x})}{\sigma(\mathbf{x})}\right) + \sigma(\mathbf{x})\,\phi\!\left(\frac{\tau - \mu(\mathbf{x})}{\sigma(\mathbf{x})}\right)$$

where $\tau = \min_i y_i$ is the current best observed value, and $\Phi$, $\phi$ are the standard normal CDF and PDF.

```{r}
#| label: ei-function

expected_improvement <- function(mu, sigma, tau, xi = 0.01) {
  imp <- tau - mu - xi
  z   <- imp / sigma
  ei  <- imp * pnorm(z) + sigma * dnorm(z)
  ei[sigma <= 0] <- 0
  ei
}
```

## Sequential Bayesian Optimisation Loop

```{r}
#| label: seq-bo-loop
#| cache: true

set.seed(123)
n0        <- 8
X_init    <- randomLHS(n0, 2)
y_init    <- mapply(goldstein_price, X_init[,1], X_init[,2])
X_current <- X_init
y_current <- y_init
history   <- list()
n_budget  <- 12

for (iter in seq_len(n_budget)) {
  gp <- tryCatch(
    km(formula  = ~1,
       design   = data.frame(x1=X_current[,1], x2=X_current[,2]),
       response = y_current,
       covtype  = "matern5_2",
       control  = list(trace=FALSE)),
    error = function(e) NULL
  )
  if (is.null(gp)) break

  pred    <- predict(gp, newdata=grid_gp[,1:2], type="UK")
  mu_g    <- pred$mean
  sd_g    <- sqrt(pred$sd2)
  tau_cur <- min(y_current)
  ei_g    <- expected_improvement(mu_g, sd_g, tau_cur)

  best_idx <- which.max(ei_g)
  x_next   <- as.numeric(grid_gp[best_idx, 1:2])
  y_next   <- goldstein_price(x_next[1], x_next[2])

  history[[iter]] <- list(
    iter    = iter,
    n_total = n0 + iter,
    X       = X_current, y = y_current,
    mu      = mu_g, sd = sd_g, ei = ei_g,
    x_next  = x_next, y_next = y_next,
    best_y  = min(y_current)
  )

  X_current <- rbind(X_current, x_next)
  y_current <- c(y_current, y_next)
}

cat("Final best value found:", round(min(y_current), 4), "\n")
cat("True minimum:          ", round(true_min_z, 4), "\n")
cat("Best point: x1 =", round(X_current[which.min(y_current),1], 3),
    "  x2 =", round(X_current[which.min(y_current),2], 3), "\n")
```

## Visualising Selected Iterations

```{r}
#| label: seq-bo-viz
#| fig-height: 14

plot_iter <- function(h) {
  df_mu <- grid_gp; df_mu$val <- h$mu
  df_ei <- grid_gp; df_ei$val <- h$ei
  X_df  <- data.frame(x1=h$X[,1], x2=h$X[,2], y=h$y)
  best_df  <- X_df[which.min(X_df$y), ]
  next_df  <- data.frame(x1=h$x_next[1], x2=h$x_next[2])

  p_mu <- ggplot(df_mu, aes(x1,x2,fill=val)) +
    geom_tile() +
    geom_point(data=X_df, aes(x1,x2), inherit.aes=FALSE,
               color="white", size=2.5, shape=16) +
    geom_point(data=best_df, aes(x1,x2), inherit.aes=FALSE,
               color="yellow", size=4, shape=8) +
    geom_point(data=next_df, aes(x1,x2), inherit.aes=FALSE,
               color="red", size=4, shape=3, stroke=2) +
    scale_fill_viridis_c(option="plasma", name="\u03bc*") +
    coord_fixed() +
    labs(title=paste0("Iter ", h$iter, " (n=", h$n_total,
                      ") \u2014 GP Mean | best=", round(h$best_y,3)),
         x=expression(x[1]), y=expression(x[2])) +
    theme_sfd + theme(legend.position="right")

  p_ei <- ggplot(df_ei, aes(x1,x2,fill=pmax(val,0))) +
    geom_tile() +
    geom_point(data=X_df,   aes(x1,x2), inherit.aes=FALSE,
               color="white", size=2.5, shape=16) +
    geom_point(data=next_df, aes(x1,x2), inherit.aes=FALSE,
               color="red", size=4, shape=3, stroke=2) +
    scale_fill_viridis_c(option="inferno", name="EI", trans="sqrt") +
    coord_fixed() +
    labs(title="Expected Improvement \u2014 next point (red cross)",
         x=expression(x[1]), y=expression(x[2])) +
    theme_sfd + theme(legend.position="right")

  p_mu / p_ei
}

iters_to_show <- c(1, 4, 8, 12)
plots_bo      <- lapply(history[iters_to_show], plot_iter)

wrap_plots(plots_bo, ncol=2) +
  plot_annotation(
    title    = "Sequential Bayesian Optimisation \u2014 Goldstein\u2013Price Function",
    subtitle = "Top: GP mean (yellow star = current best). Bottom: EI (red cross = next evaluation).",
    theme    = theme(plot.title=element_text(face="bold", size=13))
  )
```

## Convergence Curve

```{r}
#| label: convergence-curve
#| fig-height: 4.5

df_conv <- data.frame(
  iter   = 0:n_budget,
  best_y = c(min(y_init), sapply(history, function(h) min(c(h$y, h$y_next))))
)

ggplot(df_conv, aes(x=iter, y=best_y)) +
  geom_hline(yintercept=true_min_z, linetype="dashed", color="grey50", linewidth=0.8) +
  annotate("text", x=0.5, y=true_min_z+0.05, label="True minimum",
           color="grey40", size=3.5, hjust=0) +
  geom_line(color=col2, linewidth=1.2) +
  geom_point(color=col1, size=3) +
  scale_x_continuous(breaks=0:n_budget) +
  labs(
    title    = "Convergence of Bayesian Optimisation",
    subtitle = paste0("Starting from ", n0, " LHS points, ",
                      n_budget, " sequential EI-guided evaluations."),
    x="Sequential evaluation number", y="Best value found so far"
  ) + theme_sfd
```

## Comparison: Sequential EI vs Pure Space-Filling

```{r}
#| label: seq-vs-spacefilling
#| fig-height: 4.5

set.seed(123)
n_total_budget <- n0 + n_budget
X_lhs_full     <- randomLHS(n_total_budget, 2)
y_lhs_full     <- mapply(goldstein_price, X_lhs_full[,1], X_lhs_full[,2])

set.seed(99)
ord      <- sample(n_total_budget)
best_lhs <- cummin(y_lhs_full[ord])
best_seq <- df_conv$best_y[-1]

df_compare_conv <- bind_rows(
  data.frame(eval=seq_len(n_total_budget), best=best_lhs,
             method="LHS (space-filling)"),
  data.frame(eval=seq_len(n_budget) + n0,
             best=best_seq[seq_len(min(n_budget, n_total_budget-n0))],
             method="Sequential EI")
)

ggplot(df_compare_conv, aes(x=eval, y=best, color=method)) +
  geom_hline(yintercept=true_min_z, linetype="dashed", color="grey60") +
  geom_line(linewidth=1.2) +
  geom_point(size=2.5) +
  scale_color_manual(values=c("LHS (space-filling)"=col1, "Sequential EI"=col2)) +
  labs(
    title    = "Space-Filling LHS vs Sequential EI \u2014 Same Total Budget",
    subtitle = paste0("Total budget = ", n_total_budget,
                      " evaluations. Sequential EI exploits GP uncertainty to focus effort."),
    x="Total evaluations", y="Best f(x) found", color=NULL
  ) + theme_sfd
```

------------------------------------------------------------------------

# Summary

## Design Choice Guide

| Situation | Recommended Design |
|----|----|
| No prior knowledge, moderate $n$, $d \leq 10$ | **Optimised LHS** (`optimumLHS`) |
| Need to add points later | **Sobol sequence** (nested property) |
| $d > 10$, large $n$ | **Sobol** or **Halton** (low discrepancy) |
| Strict coverage guarantee needed | **Maximin** or **PhiP** design |
| Goal-directed (optimise, find contour) | **Sequential design + EI / UCB** |
| Irregular domain | Rejection sampling + **Maximin optimisation** |

## Key Takeaways

1.  **Random sampling is rarely optimal** for computer experiments: it wastes budget on clusters.
2.  **LHS** guarantees uniform 1D projections with modest computation — the go-to starting point.
3.  **Sobol and Halton** achieve provably low discrepancy and scale gracefully to higher $n$.
4.  **Maximin / Minimax** designs give explicit coverage guarantees at the cost of higher computation.
5.  **Sequential design with EI** is the most efficient strategy when you have a specific optimisation goal: it focuses evaluations where they matter most, converging faster than any static design.
6.  For **non-rectangular domains**, rejection sampling combined with maximin optimisation provides a practical and flexible solution.

------------------------------------------------------------------------

*Article written in R with Quarto, using `lhs`, `randtoolbox`, `DiceDesign`, `DiceKriging`, `ggplot2`, and `patchwork`.*
